---
title: "Tasks 1+2 - Jade"
format: html
---
```{r}
library(tidyverse)
raw_data <- read_csv(here::here("data/biomarker-raw.csv"))
```


## Task 1

### 1.  What do you imagine is the reason for log-transforming the protein levels in `biomarker-raw.csv`? (Hint: look at the distribution of raw values for a sample of proteins.)

```{r}
raw_data_long <- raw_data |> 
  select(Group:`Induced myeloid leukemia cell differentiation protein Mcl-1`, -`Target Full Name`) |> 
  mutate(ID = row_number()) |> 
  pivot_longer(`E3 ubiquitin-protein ligase CHIP`:`Induced myeloid leukemia cell differentiation protein Mcl-1`) |> 
  filter(ID > 1) |> 
  mutate(value = as.numeric(value))

raw_data_long |> 
  filter(value <= 20000) |> 
  ggplot(aes(x = value)) + 
  geom_histogram() + 
  facet_wrap(~name)

raw_data_long |> 
  group_by(name) |> 
  summarize("Mean Level" = mean(na.omit(value)))
```
We see that the protein levels in `biomarker-raw.csv` seem to be incredibly high, with a lot of the proteins having mean levels at or above 1000.


## Task 2

### 2.  Temporarily remove the outlier trimming from preprocessing and do some exploratory analysis of outlying values. Are there specific *subjects* (not values) that seem to be outliers? If so, are outliers more frequent in one group or the other? (Hint: consider tabluating the number of outlying values per subject.)
