---
title: "Biomarkers of ASD"
subtitle: "If you want a subtitle put it here"
author: "List names here"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

```{r}
# load any other packages and read data here
library(tidyverse)
library(infer)
library(randomForest)
library(tidymodels)
library(modelr)
library(yardstick)
library(glmnet)
library(ggplot2)
load('../data/biomarker-clean.RData')
raw_data <- read_csv('../data/biomarker-raw.csv')
```

## Abstract

<!-- Write a brief one-paragraph abstract that describes the contents of your write-up. -->

## Dataset

Data was collected on 154 male pediatric subjects, 76 of which were diagnosed with autism spectrum disorder (ASD) and the other 78 were determined to be typically developing (TD). Demographic data was collected for all participants, including age, ethnicity, co-morbidities, and psychiatric medications. The mean age of the ASD group was 5.6 years old, and the mean age of subjects from the TD group was 5.7 years old. Additionally, the SOMAScan platform 1.3k was used to measure the levels of 1,317 proteins in each subject. The levels of these proteins were then log10-transformed, standardized, and then outliers were trimmed to values of -3 or 3 depending on their sign.

## Summary of published analysis

<!-- Summarize the methodology of the paper in 1-3 paragraphs. You need not explain the methods in depth as we did in class; just indicate what methods were used and how they were combined. If possible, include a diagram that depicts the methodological design. (Quarto has support for GraphViz and Mermaid flowcharts.) Provide key results: the proteins selected for the classifier and the estimated accuracy. -->

## Findings

### Impact of preprocessing and outliers

#### Log-Transform

```{r}
raw_data_long <- raw_data |> 
  select(Group:`Induced myeloid leukemia cell differentiation protein Mcl-1`, -`Target Full Name`) |> 
  mutate(ID = row_number()) |> 
  pivot_longer(`E3 ubiquitin-protein ligase CHIP`:`Induced myeloid leukemia cell differentiation protein Mcl-1`) |> 
  filter(ID > 1) |> 
  mutate(value = as.numeric(value))

raw_data_long |> 
  filter(value <= 20000) |> 
  ggplot(aes(x = value)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~name) +
  labs(title = "Original Distributions of Protein Levels", y = "Count", x = 'Protein Level')

raw_data_long |> 
  group_by(name) |> 
  summarize("Mean Level" = mean(na.omit(value))) |> 
  kableExtra::kable(
    caption = "Mean Levels of Selected Proteins",
    col.names = c("Protein", "Mean Level") 
  ) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F,
    font_size = 14
  )
```

We see that the protein levels in `biomarker-raw.csv` seem to be incredibly high, with a lot of the proteins having mean levels at or above 1000. Additionally, we see hat a lot of the distributions are pretty heavily right-skewed, and thus are likely not to be normally distributed.

In contrast, when we look at the same proteins with log-transformed levels below, we see that the now the average protein level is close to 3, which is significantly lower than it was. Also, we see that the log-transform has decreased the skew of the distribution of the protein levels, causing them to be more similar to a normal distribution. We want to work with normally distributed data as often as possible, as it is easy to work with, and many data analytic techniques require normally distributed data.

```{r}
raw_data_long_log <- raw_data_long |> 
  mutate(log_value = log10(value))

raw_data_long_log |> 
  ggplot(aes(x = log_value)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~name) +
  labs(title = "Distributions of Protein Levels After Log-Transform", y = "Count", x = 'Log Protein Level')

raw_data_long_log |> 
  group_by(name) |> 
  summarize("Mean Log Level" = mean(na.omit(log_value))) |> 
  kableExtra::kable(
    caption = "Mean Log Levels of Selected Proteins",
    col.names = c("Protein", "Mean Log Level") 
  ) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F,
    font_size = 14
  )

```

#### Examining Outliers

```{r}
# get names
var_names <- read_csv(here::here('data/biomarker-raw.csv'), 
                     col_names = F, 
                     n_max = 2, 
                     col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()

# function for trimming outliers (good idea??)
trim <- function(x, .at){
  x[abs(x) > .at] <- sign(x[abs(x) > .at])*.at
  return(x)
}

# read in data
biomarker_clean_outliers <- read_csv(here::here('data/biomarker-raw.csv'), 
         skip = 2,
         col_select = -2L,
         col_names = c('group', 
                       'empty',
                       pull(var_names, abbreviation),
                       'ados'),
         na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # log transform, center and scale, and trim
  mutate(across(.cols = -c(group, ados), 
                ~ (scale(log10(.x))[, 1]))) %>%
  # reorder columns
  select(group, ados, everything())

```

#### 

```{r}
longer_outliers <- biomarker_clean_outliers |> 
  mutate(ID = row_number()) |> 
  pivot_longer(CHIP:PLXB2)

longer_outliers |> 
  filter(value <= -3 | value >= 3) |> 
  group_by(name) |> 
  summarize(n = n()) |> 
  mutate(Group = case_when(
    n <= 2 ~ "0-2 subject with an outlier value",
    n > 2 & n <= 5 ~ "3-5 subjects with outlier values",
    TRUE ~ ">5 subjects with outlier values"
  )) |> 
  group_by(Group) |> 
  summarise('Number of Proteins' = n()) |> 
  kableExtra::kable(
    caption = "Grouping Proteins by Number of Subjects with Outlier Value",
    col.names = c("Protein Group", "Number of Proteins") 
  ) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F,
    font_size = 14
  )
```

Most of the proteins had an outlier value in at least one participant, only `r 9 / 1317 * 100`% of proteins have 6 or more participants with an outlier value. The majority of proteins do not have have multiple subjects with an outlier value.

#### Are certain subjects more likely to have outlier values?

```{r}
outliers_per_id <- longer_outliers |> 
  filter(value <= -3 | value >= 3) |> 
  group_by(ID) |> 
  summarise('Number of Protiens with Outlier Values' = n(),
            group = group[1]) |> 
    mutate('Outlier Grouping' = case_when(
    `Number of Protiens with Outlier Values` <= 40 ~ "Less than or equal to 40 outlier proteins",
    `Number of Protiens with Outlier Values` > 40 & `Number of Protiens with Outlier Values` <= 100 ~ "40-100 outlier proteins",
    TRUE ~ " >100 outlier proteins"
  ))

outliers_per_id |> 
  group_by(`Outlier Grouping`) |> 
  summarize('Number of Subjects' = n()) |> 
  kableExtra::kable(
    caption = "Distribution of Amount of Outlier Proteins per Subjects",
    col.names = c("Outlier Grouping", "Number of Subjects") 
  ) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F,
    font_size = 14
  )

outliers_per_id |> 
  filter(`Number of Protiens with Outlier Values` >= 40) |> 
  kableExtra::kable(
    caption = "Outlier Subjects",
    col.names = c("ID", "Number of Proteins with Outlier Values", "Group", "Outlier Grouping") 
  ) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F,
    font_size = 14
  )
```

We see the majority of subjects seem to have less than or equal to 40 outlier proteins, with only 11 subjects having more than 40 proteins with outlier levels.

There do seem to be particular subjects that consistently have outlier protein values, with some subjects having outlier values in over 100 different proteins. Such subjects seem more likely to be in the typical developing group, with 8 subjects from the typical developing group having 40 or more proteins with outlier levels, and only 3 subjects from the ASD group having 40 or more proteins with outlier levels.

### Methodlogical variations

Task 3

### Improved classifier

```{r}
## Lasso Regression
###################

# Create class category
biomarker <- biomarker_clean %>% 
  select(-ados) %>% 
  mutate(class = as.numeric(group == 'ASD'))

# Partition
set.seed(102824)
partitions <- biomarker %>% 
  initial_split(prop=0.8)

x_train <- training(partitions) %>% 
  select(-group, -class) %>% 
  as.matrix()

y_train <- training(partitions) %>% 
  pull(class)

# Lambda Selection
cv_out <- cv.glmnet(x_train,
                    y_train,
                    family = 'binomial',
                    alpha = 1)
cv_out_df <- tidy(cv_out)

# Fit best lambda
best_lambda <- cv_out_df %>% 
  arrange(lambda) %>% 
  filter(nzero<7) %>% 
  slice(1) %>% 
  pull(lambda)

final_lasso <- glmnet(x_train,
                      y_train,
                      family = 'binomial',
                      alpha=1,
                      lambda = best_lambda)
lasso_coefs <- coef(final_lasso)
lasso_coefs_df <- as.data.frame(as.matrix(lasso_coefs)) %>% 
  filter(s0!=0)

proteins_lasso <- lasso_coefs_df %>% 
  mutate(proteins = rownames(lasso_coefs_df)) %>% 
  filter(rownames(lasso_coefs_df) != "intercept") %>%
  slice(-1) %>% 
  pull(proteins)

## Logistic Regression
#######################

# Pull proteins
fit_train <- training(partitions) %>% 
  select(group, any_of(proteins_lasso)) %>% 
  mutate(class = (group == 'ASD')) %>% 
  select(-group)
  
fit_test <- testing(partitions) %>% 
  select(group, any_of(proteins_lasso)) %>% 
  mutate(class = (group == 'ASD')) %>% 
  select(-group)

# Fit model with training data
fit <- glm(class ~ .,
           data = fit_train,
           family = 'binomial')

# Evaluate errors on test set
class_metrics <- metric_set(sensitivity, 
                            specificity, 
                            accuracy,
                            roc_auc)

fit_test %>%
  add_predictions(fit, type = 'response') %>%
  mutate(est = as.factor(pred > 0.5), tr_c = as.factor(class)) %>%
  class_metrics(estimate = est,
                truth = tr_c, pred,
                event_level = 'second')

```
